
\documentclass[12pt]{article}
\setlength{\oddsidemargin}{27mm}
\setlength{\evensidemargin}{27mm}
\setlength{\hoffset}{-1in}
\usepackage{minted}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{ stmaryrd }
\usepackage{bussproofs}
\usepackage{tikz-cd}

\newenvironment{bprooftree}
  {\leavevmode\hbox\bgroup}
  {\DisplayProof\egroup}


\newtheorem{theorem}{Theorem}


\newcommand{\hask}{\mintinline{haskell}}
\newcommand{\llb}{\llbracket}
\newcommand{\rrb}{\rrbracket}

\setlength{\topmargin}{27mm}
\setlength{\voffset}{-1in}
\setlength{\headheight}{0pt}
\setlength{\headsep}{0pt}

\setlength{\textheight}{235mm}
\setlength{\textwidth}{155mm}

%\pagestyle{empty}
\pagestyle{plain}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\renewcommand{\labelitemi}{$\diamond$}

\begin{document}
\baselineskip 12pt

\begin{center}
\textbf{\Large Type Level Shenanigans} \\
\end{center}

\section{ What Is \hask{List}}

In haskell, we often talk about \hask{List} or \hask{Maybe}.
But in system F it's not entirely clear how to encode this.
We talked about encoding data types in System F using the church encoding last time.
But when we think about 
\begin{minted}{haskell}
data Mabye t = Just t | Nothing
\end{minted}
We haven't really talked about what to do with that \hask{t}.
It should, more-or-less look like
\[
 \forall \alpha, (t \rightarrow \alpha) \rightarrow \alpha \rightarrow \alpha
\]
but \hask{t} isn't a type.
And we don't really have a sensible notion of an ``open type".
One natural choice is to put a quantifier around it.
\[
  \forall \beta, \forall \alpha, (\beta \rightarrow \alpha) \rightarrow \alpha \rightarrow \alpha
\]
This is an OK choice, but it's a bit fishy.
In particular we also talk about Maybe as it applies to thing. Such as \hask{Maybe Unit}.
However, \textbf{there is not type level application in system F}, so we can't write 
\[
  \big(\forall \beta, \forall \alpha, (\beta \rightarrow \alpha) \rightarrow \alpha \rightarrow \alpha\big) \big(\forall \alpha, \alpha \rightarrow \alpha\big)
\]

However, we can write \hask{Maybe\, Unit} directly, as 
\[
  \forall \alpha, (\big(\forall \alpha, \alpha \rightarrow \alpha\big) \rightarrow \alpha) \rightarrow \alpha \rightarrow \alpha\big)
\]
But that's not really the type \hask{Maybe} applied to the type \hask{Unit}.
It's a weird abomination, but you know what I mean.
Let's start by explaining to System F what we mean.

\section{$F_1$}

We will add two features to our language.
It may shock you to learn that it's going to be a pair of abstraction and application.
First is the abstraction. 
The concept is called a universe.
The universes are going to be tiered.
We will start with the 0, which is just the set of system F types.
We introduce a helper concept called a kind, and say they all have kind $*$. 
Then we add a new universe, and a new kind $* \rightarrow *$. 
These are the set of terms 
\[
  \alpha \Rightarrow t
\]
such that $\forall \alpha, t$ is kind $*$. 
In other words $t$ is an open term with one free variable $\alpha$.

The second thing is a type-level application, so given $\alpha \Rightarrow T : * \rightarrow *$, and a $t : *$, we define $T\, a$ as $T [\alpha := t]$. 

This already lets us give meaning to $\hask{Maybe}$.
We can now define a kind $* \rightarrow *$ as
\[
  Maybe := \beta \Rightarrow \forall \alpha, (\beta \rightarrow \alpha) \rightarrow \alpha \rightarrow \alpha
\]
In haskell terms, this setup eta expands everything, so $\hask{Maybe}$ is just shorthand for $\alpha \Rightarrow \hask{Maybe}\, \alpha$.
Of course haskell doesn't have type-level lambdas, so this is all a bit annoying to notate.

But now we can do all sorts of fun things.
We can define functors, monads, traversable, and that whole family of behaviors.
But we can't define monad transformers.
A monad transformer takes a monad, and returns a monad.
In other words it takes a $* \rightarrow *$ and returns a $* \rightarrow *$.
This is illegal since 
\begin{enumerate}
    \item there are only two universes in $F_1$, just $*$ and $* \rightarrow *$.
    \item Certainly this isn't a member of $*$, since it contains parts of $F_1$.
    \item it can't be in $F_1$, since members of $F_1$ can only take members of $*$ as arguments.
\end{enumerate}
So we need to generalize to $F_2$.
Let's do this \textit{to exhaustion}.

\section{$F_\omega$}
The idea is to define $F_{n+1}$ as follows. The universe $n+1$ is the term terms\[
  \alpha \Rightarrow t
\]
such that for any $k$ in universe $\leq n$, $t[\alpha := k]$ is in universe $n$.
So now we can build $(* \rightarrow *) \rightarrow (* \rightarrow *)$, which is the type of monad transformers.
In fact, we can build all sorts fun things. Note that the distinction between kinds and universes shows up here, since 
\[
  * \rightarrow (* \rightarrow *) \quad\text{and}\quad (* \rightarrow *) \rightarrow (* \rightarrow *)
\]
are both in universe 2. 
Then we define higher-kinded type application the same way. 
Given kinds $k_1$ and $k_2$, and types $t : k_1 \rightarrow k_2$ and $t' : k_1$, we define $t\, t'$ as $t[\alpha := t']$.
And then a simple induction proves that this has kind $k_2$.

So finally, we can define $F_\omega$ as the the union of $F_n$. 
Or, more formally for any category theorist, we observe that $F_n$ embeds in $F_{n+1}$.
So we can take a categorical limit.

The upshot is we can now define \textit{any} type level operator we want, with parameters coming from anywhere. But we have to be careful. What happens to type-checking? What happens to inference? Is this confluent, normalizing, etc?
\begin{enumerate}
    \item Hindle-Milner$_\omega$ type checks the same way. Basically any term only has finitely many universes, so it's not that bad.
    \item Of course things normalize, we haven't changed any terms! All of the terms of System F have types whose kind is $*$.
    \item There are still theorems for free. Since all terms of of type $*$ this isn't interesting.
\end{enumerate}

\section{Church Encodings in $F_\omega$}
Last time we saw church encodings for $*$ kinded types with no recursion. 
This generalizes easily to $k_1 \rightarrow *$, like we saw with \hask{Maybe}.

But we can now finally address the question of recursive ADTs.

The first thing we do is \textbf{continuation passing transform at the type level}.
That is, we will first perform a ``haskell-to-haskell" data transform.
Add an extra type parameter, and replace every recursive call with that parameter.
\[
 \hask{Nat = S Nat | Zero} \quad\longrightarrow\quad \hask{NatK r = S r | Zero}
\]
Now we can encode \hask{NatK} the standard way 
\[
  NatK := r \Rightarrow \forall \alpha, (r \rightarrow \alpha)  \rightarrow \alpha \rightarrow  \alpha
\]
But this is either zero, or one plus a continuation. 
If we wanted to represent two, we need

\[
  NatK \circ NatK := r \Rightarrow \forall \alpha, (\Big( \forall \beta, (r \rightarrow \beta)  \rightarrow \beta \rightarrow  \beta\Big) \rightarrow \alpha)  \rightarrow \alpha \rightarrow  \alpha
\]

And we want a fixpoint of these types. 
How can we do that?
Well, there is no one best way.
But a very reasonable choice is to consider finite chunks of the fixpoint. 
So we want to think about $(NatK \circ ... \circ NatK)\, Unit$, which precisely the naturals up to $n$.
Then we want some type which has all of these in it.
In other words we want a type $Nat$ where there exists functions
\begin{equation}
\begin{split}
 &f_0 : Unit \rightarrow Nat\\
 &f_1 : NatK\; Unit \rightarrow Nat\\
 &f_2 : NatK\; (NatK\; Unit) \rightarrow Nat \\
 & ...
\end{split}
\end{equation}
which respect the truncation $NatK (NatK\, Unit) \rightarrow NatK\, Unit$.
And where each $f_1$ is an injection.
There is more than one way to do this, this is the difference between the Scott and the Church encoding. 
We will look at the church encoding, which is simply 
\[
\forall \alpha, (\alpha \rightarrow \alpha) \rightarrow \alpha \rightarrow \alpha
\]

So long story short, we replace $r$ with $\alpha$.
Note that this has the desired property, and for example

\[
  f_0 := \Lambda \alpha. \lambda (s : \alpha \rightarrow \alpha). \lambda (z: \alpha). z
\]
\[
  f_1 := \lambda (f : NatK\; Unit). \Lambda \alpha. \lambda (s : \alpha \rightarrow \alpha). \lambda (z: \alpha). f \;\alpha\; \Big(f_0\,  id\,\alpha\, s\, (s\, z) \Big)\; z 
\]
and
\begin{equation}
\begin{split}
f_2 := &\lambda (f : NatK(NatK\; Unit)). \Lambda \alpha. \lambda (s : \alpha \rightarrow \alpha). \lambda (z: \alpha). \\
 & f \;\alpha\; \Big(\lambda (n : NatK\, Unit). f_1\, n\, \alpha\, s\, (s\, z)\Big) z
\end{split}
\end{equation}
This is basically just calling the previous one with an extra $s$.
The interesting thing is that the result from before holds.
Where the encoding of $Nat$ is equivalent to the one you'd get extending System F with explicit Nat terms.
But now the elimination rule is a bit trickier, since you don't want to represent arbitrary recursion. 
As the haskell program 
\begin{minted}{haskell}
f x = case x of 
        Zero -> f Zero
        S n -> f (S n)
\end{minted}
does not have a corresponding $F_\omega$ term since it doesn't terminate.
But the upshot of this is that we get a legit System F only description of Nat.
Where all of our nice recursive functions work.
In particular, recursion structured like 
\begin{minted}{haskell}
f :: Nat -> t
f x = case x of 
        Zero -> z
        S n -> let recursed := f n in s recursed
\end{minted}
for some \hask{z :: t} and \hask{s :: t -> t}.
But this process will works for any recursive ADT.
The construction of the $f_i$ is a bit annoying, but it will work!

\section{Folding over anything}
This leads to a series of important features in system $F_\omega$.
Armed with
\[
  Nat := \forall \alpha, (\alpha \rightarrow \alpha) \rightarrow \alpha \rightarrow \alpha
\]
As our encoding of
\[
\hask{data Nat = S nat | Zero}
\]
system F now knows about arithmetic.
In fact for any of these structures, the church encodings define ``foreach loops" that respect the structure of the data.
They are guaranteed to terminate. 
So we have basic loops, but also termination guarantees.
This leads to the rather critical code design point: don't use general recursive when a church encoding will suffice.  This leads to a series of interesting design patterns called ``bananas, envelopes, and barbed wires".
We will look at that paper next week.
But the idea is that the theorems for free for these encodings will allow us to fuse to constructors and destructors for a huge performance gain. 

We can also now talk about the haskell type \hask{Functor}.
Given a $t : * \rightarrow *$, we equip it with an 
\[
  fmap : \forall \alpha, \forall \beta, (\alpha \rightarrow \beta) \rightarrow t \alpha \rightarrow t\beta
\]

Along with laws that 
\[
   fmap\; id = id\quad\text{and}\quad fmap (f \circ g) = fmap\; f \circ fmap\; g
\]
Oddly, these are exactly the theorems for free for $t$.
Something deep is going on, and is one of the reasons why deriving functors is so safe.
It's just deriving a property of your data type directly from System F.
\section{A Moment of Reflection}
We have now pretty much defined the core of the ML/Haskell languages.
We've build the main objects of study in lambda calculus, although many more extensions are possible and interesting. 
And while we often talk about trading off expressiveness for analyzability, we have broken that down along several dimensions
\begin{enumerate}
    \item Confluence (I want evaluation to have a unique meaning)
    \item Normalization (I want things to terminate)
    \item Loops (I want to iterate over things)
    \item Term Polymorphism (I want only one identity function)
    \item Type Polymorism AKA Universes (I want only one List type)
    \item Impredicativity (I want a type with a ``forall types" in it)
    \item Principal types (I want to say "the type" of a term, not "a type")
    \item Type computations (I want the computer to tell me "the" type)
    \item Recursion (I want to work with any nat, not just fixed sizes ones)
\end{enumerate}
The idea is that we can make more careful judgements about what kind of expresivity is needed, and where the tradeoffs can be made when designing a language.

\end{document}

